Spitzer, Gaia and the Potential of the Milky Way

### 2013-07-24 text/paper1/paper.tex
change _orbit to just R

Most recently, \citet[][hereafter LM10]{law2010} have combined all
current data on the Sgr debris to assess the triaxiality and orientation
of the outer Galaxy, the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible. Figure \ref{fig:lm10} shows particle positions from the final time-step of an N-body simulation of dwarf satellite disruption along the expected Sgr
orbit in the best-fitting Milky Way halo model from LM10. [TODO: Maybe not true...] The positions are shown in a coordinate system defined
by the orbit of the satellite \citep{majewski2004}. The
simulation was run in a three-component potential, with a triaxial,
logarithmic halo model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2)
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis ratios ($q_1$, $q_2$) and orientation of the halo with respect to the baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}

@@

\subsection{The algorithm: Rewinder}
Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_{i}, \bs{v}_{i})|_{t=0}$ (e.g., measurements of all position and velocity components for these stars \emph{today} at $t=0$), the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi({\boldsymbol\theta})$. For a
given set of potential parameters, $\boldsymbol\theta$, the orbits of the stars and
progenitor are integrated backwards for several Gigayears (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become physically re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates are
computed 
\begin{equation}
  \bs{q}_{i} = \frac{\bs{x}_{i} -
    \bs{x}_{p}}{R_{\rm tide}}\,\,\,\,;\,\,\,\,\bs{p}_{i} = \frac{\bs{v}_{i} -
    \bs{v}_{p}}{v_{\rm esc}}
\end{equation}
where $(\bs{x},\bs{v})_{i}$ and $(\bs{x},\bs{v})_{p}$ are the phase-space coordinates for the particles and progenitor, respectively. These definitions require an estimate of the mass of the
satellite, $m_{sat}$, which, combined with the computed enclosed mass
of the potential, $m_{enc}$, sets the instantaneous tidal radius and
escape velocity,
\begin{equation}
  R_{\rm tide}=\Big(\frac{m_{\rm sat}}{3m_{\rm enc}}\Big)^{1/3}\,\,\,\,\,\,\,\,
  v_{\rm esc}=\sqrt{\frac{2Gm_{\rm sat}}{R_{\rm tide}}}.
\end{equation}

Qualitatively, when the distance in this normalized 6D space,
$D_{\rm ps}=\sqrt{|\bs{q}_{i}|^2+|\bs{p}_{i}|^2} < \sqrt{2}$, the star is
probably recaptured by the satellite, but this is not imposed as a
hard constraint. [JZSH ...] As a statistical measure of how many stars have recombined, we
construct the 6D distribution of the minimum phase-space vectors for
each star over each individual, integrated orbit. For each star, $i$,
the phase-space distance, $D_{\rm ps}$, is computed at each timestep
$t_{j}$, and the vector with the minimum phase-space distance is stored
\begin{align}
  t^*_{i} &= \argmin_{t} D_{{\rm ps},i}\\
  \bs{A}_{i} &= (\bs{q}_{i}(t^*_{i}),\bs{p}_{i}(t^*_{i})).
\end{align}
Thus, the matrix $\bs{A}_{ik}$ contains these minimum phase-space distance vectors for each star,
where $k\in\{1,6\}$. Intuitively, the variance of this distribution of minimum distance vectors
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars; in an incorrect potential, small differences in the
orbits of the stars relative to the orbit of the progenitor cause them
to spread out in this relative, normalized phase space. Thus, the
\emph{generalized variance} of the distribution, computed for a given
set of potential parameters, $\bs{\theta}$, is a natural choice for
the scalar objective function used in constraining the potential of
the Milky Way,
\begin{equation}
  f(\bs{\theta})=\det(\mathrm{Cov}(A_{ik})).
\end{equation}
[This objective function is an approximation to a more general likelihood function, described in detail in forthcoming work (Price-Whelan \& Johnston in prep.). The results shown in sections xx and yy are meant to illustrate the strength and concept of the method..]

>>>>

Most recently, \citet[][hereafter LM10]{law2010} have combined all
current data on the Sgr debris to assess the triaxiality and orientation
of the outer Galaxy, the first time that such a reconstruction of
the 3-dimensional mass distribution of a dark matter halo has been
feasible. Figure \ref{fig:lm10} shows particle positions from the final time-step of an N-body simulation of dwarf satellite disruption along the expected Sgr
orbit in the best-fitting Milky Way halo model from LM10. [TODO: Maybe not true...: The positions are shown in a coordinate system defined
by the orbit of the satellite \citep{majewski2004}]. The
simulation was run in a three-component potential, with a triaxial,
logarithmic halo model of the form
\begin{equation}
  \Phi_{halo} = v_{halo}^2 \ln(C_1 x^2 + C_2 y^2 + C_3 xy + (z/q_z)^2 + R_c^2)
\end{equation}
where $C_1$, $C_2$, and $C_3$ are combinations of the $x$ and $y$ axis ratios ($q_1$, $q_2$) and orientation of the halo with respect to the baryonic disk ($\phi$):
\begin{align}
  C_1 &= \frac{\cos^2\phi}{q_1^2} + \frac{\sin^2\phi}{q_2^2}\\
  C_2 &= \frac{\sin^2\phi}{q_1^2} + \frac{\cos^2\phi}{q_2^2}\\
  C_3 &= 2\sin\phi\cos\phi \left(q_1^{-2} - q_2^{-2}\right).
\end{align}

@@

\subsection{The algorithm: Rewinder}
Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_{i}, \bs{v}_{i})|_{t=0}$ (e.g., measurements of all position and velocity components for these stars \emph{today} at $t=0$), the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi({\boldsymbol\theta})$. For a
given set of potential parameters, $\boldsymbol\theta$, the orbits of the stars and
progenitor are integrated backwards for several Gigayears (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become physically re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates are
computed 
\begin{equation}
  \bs{q}_{i} = \frac{\bs{x}_{i} -
    \bs{x}_{p}}{R_{\rm tide}}\,\,\,\,;\,\,\,\,\bs{p}_{i} = \frac{\bs{v}_{i} -
    \bs{v}_{p}}{v_{\rm esc}}
\end{equation}
where $(\bs{x},\bs{v})_{i}$ and $(\bs{x},\bs{v})_{p}$ are the phase-space coordinates for the particles and progenitor, respectively. These definitions require an estimate of the mass of the
satellite, $m_{sat}$, which, combined with the computed enclosed mass
of the potential, $m_{enc}$, and the orbital radius of the satellite, $R$, sets the instantaneous tidal radius and
escape velocity,
\begin{equation}
  R_{\rm tide}=R\Big(\frac{m_{\rm sat}}{3m_{\rm enc}}\Big)^{1/3}\,\,\,\,;\,\,\,\,
  v_{\rm esc}=\sqrt{\frac{2Gm_{\rm sat}}{R_{\rm tide}}}.
\end{equation}

Qualitatively, when the distance in this normalized 6D space,
$D_{{\rm ps},i}=\sqrt{|\bs{q}_{i}|^2+|\bs{p}_{i}|^2} < \sqrt{2}$, the star is
likely recaptured by the satellite, but this is not imposed as a
hard constraint. [JZSH ...] As a statistical measure of how many stars have recombined, we
construct the 6D distribution of the minimum phase-space vectors for
each star over each individual, integrated orbit. For each star, $i$,
the phase-space distance, $D_{\rm ps}$, is computed at each timestep
$t_{j}$, and the vector with the minimum phase-space distance is stored
\begin{align}
  t^*_{i} &= \argmin_{t} D_{{\rm ps},i}\\
  \bs{A}_{i} &= (\bs{q}_{i}(t^*_{i}),\bs{p}_{i}(t^*_{i})).
\end{align}
Thus, the matrix $\bs{A}_{ik}$ contains these minimum phase-space distance vectors for each star,
where $k\in[1,6]$. Intuitively, the variance of this distribution of minimum distance vectors
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars; in an incorrect potential, small differences in the
orbits of the stars relative to the orbit of the progenitor cause them
to spread out in this relative, normalized phase space. Thus, the
\emph{generalized variance} of the distribution, computed for a given
set of potential parameters, $\bs{\theta}$, is a natural choice for
the scalar objective function used in constraining the potential of
the Milky Way,
\begin{equation}
  f(\bs{\theta})=\det(\mathrm{Cov}(A_{ik})).
\end{equation}
[This objective function is an approximation to a more general likelihood function, described in detail in forthcoming work (Price-Whelan \& Johnston in prep.). The results shown in sections xx and yy are meant to illustrate the strength and concept of the method..]
